---
title: "04_miRNA_analysis: KEGG pathway search"
author: "D. Ryazantsev, E. Sharova"
date: "`r Sys.Date()`"
output:
  github_document: 
    toc: yes
    #toc_float: yes
    toc_depth: 5
    number_sections: true
  pdf_document: 
    latex_engine: xelatex
    includes:
      in_header: preamble.tex
  word_document: 
#bibliography: bibliography.bib
#csl: gost-r-7-0-5-2008-numeric.csl
link-citations: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, echo=F, error=F, message=FALSE, warning=FALSE}
library(devtools)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(glue) # f-strings analog
library(pkgconfig)
library(tximport)
library(GenomicFeatures)
library(DT)
library(gridExtra)
library(grid)
library(matrixStats)
library(limma) # getGeneKEGGLinks function there
library(UpSetR)
library(BiasedUrn)
library(ComplexHeatmap)
library(pathview)
library(dplyr)
library(DOSE)
# renv install worked after this installation: 
#package_url <- "https://cran.r-project.org/src/contrib/Archive/BH/BH_1.84.0-0.tar.gz"
#utils::install.packages(pkgs = package_url, repos = NULL)
# BiocManager::install("fgsea")
options(bitmapType="cairo") 
```

# Analyze gene presence in KEGG pathways

To assess the potential regulatory role of target genes, we examine their presence in KEGG pathway maps.

**What we need to find:**

1. Genes present in many pathways — the most ambiguous ones
2. Genes with the highest co-occurrence within a single pathway — members of the same pathway

`pathview` works with Entrez IDs, so gene names need to be converted.

```{r}
final_FECD_MIMAT <- readRDS("/data7a/bio/human_genomics/fuchs_dystrophy/nanostring/analisys/RDS/final_FECD_MIMAT_v3.RDS")

downcorr_genes <- final_FECD_MIMAT$RNA %>% unique()

# from symbol -> entrez_id
entrez_ids_f <- bitr(downcorr_genes, fromType = "SYMBOL", 
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

# all kegg cards
all_cards <- getGeneKEGGLinks(species.KEGG = "hsa", convert = FALSE)
```

ENTREZID enrichment into big list.

```{r}
kegg_results_f <- enrichKEGG(gene = entrez_ids_f$ENTREZID, organism = "hsa", pvalueCutoff = 0.05)
kegg_results_f
```

### KEGG results

```{r}
# kegg_results@result[["Description"]]
kegg_fecd_all <- kegg_results_f@result

# get signiff
kegg_fecd <- kegg_fecd_all %>% dplyr::filter(p.adjust <= 0.05)

saveRDS(kegg_fecd, file="/data7a/bio/human_genomics/fuchs_dystrophy/nanostring/analisys/RDS/kegg_fecd_result.RDS")
#DT::datatable(kegg_fecd, filter = 'top', caption = "All cards")
kegg_fecd
```

### Plot p-value

We will build a horizontal barplot with adjusted p-values and pathway names to visualize the identified KEGG maps. We will also include the number of gene occurrences (n).

```{r plot_pv}
for_plot <- kegg_fecd_all %>% arrange(p.adjust) %>% slice_head(n=10) %>% dplyr::select(Description, pvalue, p.adjust, Count) %>% mutate(log10pv = -log10(p.adjust)) %>% 
	arrange(Count)


ggplot(for_plot, aes(x=reorder(Description, Count), y=Count, fill=p.adjust)) + geom_col(show.legend =T)  + coord_flip() +  
 # geom_label(aes(label = Count, hjust = 3, colour = "green")) + 
  labs(title = "Top 10 KEGG Pathways") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10), legend.position = "right",
  			axis.title.y = element_blank()) + 
  scale_fill_gradientn(colours = c("#ff0000", "#0000ff"),  
    name = "P-value Adjusted" ) 

#only pval
ggplot(for_plot, aes(x=reorder(Description, log10pv), y=log10pv, fill=log10pv)) + geom_col(show.legend =T)  + coord_flip() +  
  labs(title = "Top 10 KEGG Pathways") +
	geom_text(aes(label = Count), hjust = -0.3) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10), legend.position = "right",
  			axis.title.y = element_blank(), axis.title.x = element_blank()) + 
  scale_fill_gradientn(colours = c( "#0000ff", "#ff0000"),  
    name = expression(paste("-", log[10], " p-value adjusted")) ) +
	scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

#ggsave("./plots/kegg_top10.png",dpi=500)
```

### mir-29 kegg

Same plot for mir-29 targets.

```{r mir29_get_kegg}
mir29_all <- readRDS("/data7a/bio/human_genomics/fuchs_dystrophy/nanostring/analisys/RDS/mir29_targets.RDS")
mir29 <- mir29_all$gene %>% unique()
# from symbol -> entrez_id
mir29 <- bitr(mir29, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

mir29_kegg <- enrichKEGG(gene = mir29$ENTREZID, organism = "hsa", pvalueCutoff = 0.05)@result

mir29_kegg_sign <- mir29_kegg %>% dplyr::filter(p.adjust <= 0.05)

#DT::datatable(mir29_kegg_sign, filter = 'top', caption = "All cards")
mir29_kegg_sign
```

```{r}
for_plot <- mir29_kegg_sign %>% arrange(p.adjust) %>% slice_head(n=10) %>% dplyr::select(Description, pvalue, p.adjust, Count) %>% mutate(log10pv = -log10(p.adjust)) %>% 
	arrange(Count)

ggplot(for_plot, aes(x=reorder(Description, log10pv), y=log10pv, fill=log10pv)) + geom_col(show.legend =T)  + coord_flip() +  
  labs(title = "Top 10 hsa-miR-29 targets' Pathways") +
	geom_text(aes(label = Count), hjust = -0.3) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10), legend.position = "right",
  			axis.title.y = element_blank()) + 
  scale_fill_gradientn(colours = c( "#0000ff", "#ff0000"),  
    name = "-log10 P-value Adjusted" ) +
	scale_y_continuous(expand = expansion(mult = c(0, 0.17)))
```

Top 10 filled cards

```{r}
#DT::datatable(kegg_fecd[1:10,], filter = 'top', caption = "Top 10 cards")
kegg_fecd[1:10,]
```



We will identify genes that appear most frequently across different pathways.

```{r freq_counts_f}
kegg_merged_f <- left_join(all_cards, entrez_ids_f, by = c("GeneID" = "ENTREZID"))
frequent_counts_f <- kegg_merged_f %>% dplyr::count(SYMBOL, sort = TRUE) %>% drop_na()
frequent_counts_f <- frequent_counts_f %>%
  left_join(kegg_merged_f, by = "SYMBOL") %>%
  left_join(kegg_fecd %>% dplyr::select(ID, Description), by = c("PathwayID" = "ID"))

#DT::datatable(frequent_counts_f, filter = 'top')
frequent_counts_f
```

# Visualisation with pathview

```{r, eval=F}
cpm.symbol.group <- readRDS("/data7a/bio/human_genomics/fuchs_dystrophy/illumina/Fuchs_transcriptomes/analysis/transcript_analysis/RDS/expression/cpm.symbol.group_list.RDS")[[1]] %>% as.data.frame() %>% mutate_all(~ 2^.) %>% rownames_to_column('SYMBOL') 

# get ENTREZID for cpm
cpm_entrez <- bitr(cpm.symbol.group$SYMBOL, fromType = "SYMBOL", 
                   toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# maps id
kegg_ids <- kegg_fecd %>% pull(ID)
kegg_ids
```


Plot for 1 card:

```{r make1card, eval=F}
id_for_cpm <- kegg_fecd$geneID %>% str_split("/") %>% unlist() %>% unique()

for_card <-  cpm_entrez %>% left_join(cpm.symbol.group, by='SYMBOL') %>% 
	dplyr::filter((ENTREZID %in% id_for_cpm)) %>% na.omit() %>% 
  column_to_rownames(var="ENTREZID") %>% 
	dplyr::mutate(FC = (FECD_exp / Ctrl), logFC = log2(FC))

# cpm or FC
gene_vec <- for_card$FECD_exp  
names(gene_vec) <- rownames(for_card)  
card = 'hsa04510'

pathview(
  gene.data = gene_vec,  
  pathway.id = card,
  gene.idtype = "entrez",
  kegg.native = TRUE,
  limit = list(gene = c(min(gene_vec), max(gene_vec))),
  bins = list(gene = 20),
  species = "hsa",
  
  low = list(gene = "blue"),
  #mid = list(gene = "lightblue"),
  mid = list(gene = "pink"),
  high = list(gene = "red")
  )
```

Flot for all with coloring and scale. 

```{r all_cards_new_cycle, eval=F}
for_card <-  cpm_entrez %>% left_join(cpm.symbol.group, by='SYMBOL') %>% 
	dplyr::filter((ENTREZID %in% id_for_cpm)) %>% na.omit() %>% 
  column_to_rownames(var="ENTREZID") %>% 
	dplyr::mutate(FC = (FECD_exp / Ctrl), logFC = log2(FC))


for (card in kegg_ids) {
	selected_genes <- kegg_fecd %>% filter(ID == card) %>% dplyr::select(geneID) %>% str_split("/") %>% unlist() %>% unique()
	
	for_card2 <- for_card %>% dplyr::filter(rownames(for_card) %in% selected_genes)
gene_vec <- for_card2$FECD_exp  
names(gene_vec) <- rownames(for_card2)  
	
  pathview(
  gene.data = gene_vec,  
  pathway.id = card,
  gene.idtype = "entrez",
  kegg.native = TRUE,
  limit = list(gene = c(min(gene_vec), max(gene_vec))),
  bins = list(gene = 30),
  species = "hsa",
  
  low = list(gene = "lightblue1"),
  #mid = list(gene = "lightblue"),
  mid = list(gene = "pink"),
  high = list(gene = "red")
  )
}
```

